Immutable Source-Mapped Abstract Syntax Tree: \\ A Design Pattern for Refactoring Engine APIs \\ \today

### 2013-07-14 paper.tex
Second draft

\documentclass[prodmode]{acmlarge}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\newcommand{\ttt}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}

@@

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree: \\
A Design Pattern for Refactoring Engine APIs}
\author{JEFFREY L.\ OVERBEY \affil{Auburn University}}

@@

However, automated refactorings manipulate source code, so implementing an
automated refactoring requires a data structure that describes the structure of
soruce code.  Refactoring tools almost always use a data structure called an
\textit{abstract syntax tree} (AST).  ASTs describe source code hierarchically:
e.g., classes contain methods, methods contain statements, statements contain
expressions, etc.  ASTs are used in compilers and static analysis tools, but
refactoring tools impose more stringent requirements on ASTs---specifically,
the ASTs must be suitable for manipulating the user's source code.

@@

You are implementing a refactoring engine or a similar tool that makes
semi-automated edits to source code.  You have provided a mechanism that allows
third parties to contribute new refactorings/transformations.  You have an API
for adding, deleting, and changing source code files (so the mechanism to, say,
``delete the first 27 characters of \textit{main.c}'' is well-defined).

\section{Problem}

@@

Consider a simple Java refactoring that adds braces to the bodies of
\textit{if} statements, if they do not already exist.  For example, it would
change \\
\centerline{\ttt{if (a < b) if (c < d) System.out.println("a");
                            else System.out.println("b");}}
into \\
\centerline{\ttt{if (a < b) \{ if (c < d) \{ System.out.println("a"); \}
                               else \{ System.out.println("b"); \} \}}.}
Even this simple refactoring requires nontrivial knowledge of Java's syntax,
including how the dangling \textit{else} is matched with a corresponding
\textit{if}.

@@

\subsection{Abstract Syntax Trees (ASTs)}
\label{ss:asts}

@@

\begin{wrapfigure}{R}{3in} % r is right, R is float right
% Original image is 4.624 x 10.651 in
\includegraphics[scale=0.6]{jdt-ast.eps}
%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
%tree for the following program: \\
%\ \\
%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
%\hspace*{.2em} public int FIVE = 5; \\
%\}
%}}} \\
%\ \\
%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
%character in the source code.
%}
\caption{Visualization of an AST from Eclipse JDT.}
\label{fig:jdt-ast}
\end{wrapfigure}

@@

\subsection{Requirements of Refactoring ASTs}
\label{ss:reqs}

@@

\begin{enumerate}
\item To be useful for refactoring, an AST must accurately and
precisely model the user's source code at a very fine level of detail.
\item It must be possible to map nodes in the AST to locations in the
source code, and vice versa.
\end{enumerate}

@@

When ASTs are used in compilers, it is common to make them \textit{mutable},
meaning that nodes can easily be added, changed, or deleted from the tree.  For
example, a compiler might determine that \ttt{a = b*1} can be simplified to
\ttt{a = b}, and then change the AST accordingly.

@@

Unfortunately, while a mutable AST can provide an elegant means of manipulating
source code, that are other issues that can make mutable ASTs problematic from
an API design perspective.  Some notable issues include the following.
\begin{itemize}
\item \textit{AST nodes contain source location information.}  After the
structure of the AST is changed, what does the source location information in
each node represent: the ``old'' source location (in the original code) or the
``new'' location (after it has been modified)?  If it represents the old
location, what should the source location be when new nodes are inserted into
the tree?  If it represents the new location, then every time a change is made
to the tree, source location information must be updated for every node in the
tree, which may be expensive.
\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
Suppose a client traverses the AST, storing pointers to some ``interesting''
nodes in a collection.  Then, an ancestor of one of those nodes is removed from
the AST.  This will leave the client in a difficult-to-debug situation where
his collection contains pointers to nodes that no longer exist.
\item \textit{Semantic information may be invalidated.}  Refactorings often
need to determine, for a particular use of a variable, what declaration that
use corresponds to.  For example, in Java, if a field and a local variable were
both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
would refer to the field, while the second would refer to the local variable.
However, if the local variable declaration were removed, the second \ttt{x}
would refer to the field instead.  This can be confusing to clients, since
information collected about the program at one point in time may no longer be
true after the tree's structure changes.
\item \textit{It may be possible to mutate the AST so that it does not
represent a legal program.}  It might even be possible to mutate the AST so
that it is no longer a tree at all.  In Java, classes can contain other
classes.  What happens if a client tries to make a class an inner class of
itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
the AST's structure could try to prevent such things, but attempting to detect
and prevent every possible mistake would be difficult and potentially
expensive.
\item \textit{Mutability can be problematic in multithreaded code.}
\end{itemize}

@@

Note that the manipulation code using a mutable AST is simpler, and it more
directly expresses the programmer's intention.  Manipulating source code using
offset/length information has a major disadvantage: it leads to ugly client
code.  The client code can easily become littered with variables holding
character offsets and methods that perform intricate string manipulations.

>>>>

% TODO: Version with [draft] is for draft mode only
%\documentclass[prodmode]{acmlarge}
\documentclass[draft]{acmlarge}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{datetime} % for \today
\usepackage{listings} % for lstlisting environment
\usepackage{placeins} % for \FloatBarrier
\newcommand{\ttt}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}

@@

% Title portion
\title{Immutable Source-Mapped Abstract Syntax Tree: \\
A Design Pattern for Refactoring Engine APIs \\ \today}
\author{JEFFREY L.\ OVERBEY \affil{Auburn University}}

@@

However, automated refactorings manipulate source code, so implementing an
automated refactoring requires a data structure that describes the structure of
source code.  Refactoring tools almost always use a data structure called an
\textit{abstract syntax tree} (AST).  ASTs describe source code hierarchically:
e.g., classes contain methods, methods contain statements, statements contain
expressions, etc.  ASTs are used in compilers and static analysis tools, but
refactoring tools impose more stringent requirements---specifically, the ASTs
must be suitable for manipulating the user's source code.

@@

You are implementing a refactoring engine or a similar tool that makes
automated edits to source code.  You have provided a mechanism that allows
third parties to contribute new refactorings/transformations.  You have an API
for adding, deleting, and changing source code files (so the mechanism to, say,
``delete the first 27 characters of \textit{main.c}'' is well-defined).

\section{Problem}
\label{sec:Problem}

@@

Consider a simple Java refactoring that adds braces to the bodies of
\textit{if} statements, if they do not already exist.  For example, it would
make the following change.

\begin{center}
\vspace*{1em}
\begin{tabular}{ccc}
\begin{minipage}{2.25in}
\texttt{%
                if (a < b) \\
\hspace*{1.5em}   if (c < d) \\
\hspace*{3.0em}     System.out.println("a"); \\
\hspace*{1.5em}   else \\
\hspace*{3.0em}     System.out.println("b");} \\
 \\
\end{minipage}
& $\Longrightarrow$ &
\begin{minipage}{2.25in}
\texttt{%
                if (a < b) \textbf{\{} \\
\hspace*{1.5em}   if (c < d) \textbf{\{} \\
\hspace*{3.0em}     System.out.println("a"); \\
\hspace*{1.5em}   \textbf{\}} else \textbf{\{} \\
\hspace*{3.0em}     System.out.println("b"); \\
\hspace*{1.5em}   \textbf{\}} \\
                \textbf{\}}}
\end{minipage}
\end{tabular}
\vspace*{1em}
\end{center}

Even this simple refactoring requires nontrivial knowledge of Java's syntax,
including how the dangling \textit{else} is matched with a corresponding
\textit{if}.

@@

\subsection{Abstract Syntax Trees}
\label{ss:asts}

\begin{wrapfigure}{R}{3in} % r is right, R is float right
% Original image is 4.624 x 10.651 in
\includegraphics[scale=0.65]{jdt-ast.eps}
%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
%tree for the following program: \\
%\ \\
%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
%\hspace*{.2em} public int FIVE = 5; \\
%\}
%}}} \\
%\ \\
%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
%character in the source code.
%}
\caption{Visualization of an AST from Eclipse JDT.}
\label{fig:jdt-ast}
\end{wrapfigure}

@@

@@

\subsection{Fulfilling the Requirements of Refactoring}
\label{ss:reqs}

@@

\begin{enumerate}
\item To be useful for refactoring, an AST must accurately and precisely model
the user's source code at a very fine level of detail.  It may even be
necessary to include information about punctuation and comments in order to
refactor code appropriately~\cite{sommerlad08retaining}.
\item It must be possible to map nodes in the AST to locations in the
source code, and vice versa.
\end{enumerate}

@@

When ASTs are used in compilers, it is common to make them \textit{mutable},
meaning that nodes can easily be added, changed, or deleted from the tree.  For
example, a compiler might determine that \ttt{a~=~b*1} can be simplified to
\ttt{a~=~b}, and then change the AST accordingly.

@@

Unfortunately, while a mutable AST can provide an elegant means of manipulating
source code, that are other issues that can make mutable ASTs problematic from
an API design perspective.  Some notable issues include the following.
\begin{itemize}
\item \textit{AST nodes contain source location information.}  After the
structure of the AST is changed, what does the source location information in
each node represent: the ``old'' source location (in the original code) or the
``new'' location (after it has been modified)?  If it represents the old
location, what should the source location be when new nodes are inserted into
the tree?  If it represents the new location, then every time a change is made
to the tree, source location information must be updated for every node in the
tree, which may be expensive \dots or else a clever implementation (e.g.,
computation from relative offsets) must be used instead.
\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
Suppose a client traverses the AST, storing pointers to some ``interesting''
nodes in a collection.  Then, an ancestor of one of those nodes is removed from
the AST.  This will leave the client in a difficult-to-debug situation where
his collection contains pointers to nodes that no longer exist.
\item \textit{Semantic information may be invalidated.}  Refactorings often
need to determine, for a particular use of a variable, what declaration that
use corresponds to.  For example, in Java, if a field and a local variable were
both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
would refer to the field, while the second would refer to the local variable.
However, if the local variable declaration were removed, the second \ttt{x}
would refer to the field instead.  This can be confusing to clients, since
information collected about the program at one point in time may no longer be
true after the tree's structure changes.
\item \textit{It may be possible to mutate the AST so that it does not
represent a legal program.}  It might even be possible to mutate the AST so
that it is no longer a tree at all.  In Java, classes can contain other
classes.  What happens if a client tries to make a class an inner class of
itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
the AST's structure could try to prevent such things, but attempting to detect
and prevent every possible mistake would be difficult and potentially
expensive.
\item \textit{Mutability can be problematic in multithreaded code.}
\end{itemize}

@@

The situation gets a bit worse when changes are made to nested constructs.
Recall the example from Section~\ref{sec:Problem}, which added braces to two
\textit{if} statements, one nested inside the other.  Suppose the inner
\textit{if} statement were modified first: Adding braces to it would change the
position at which the closing brace for the outer \textit{if} statement would
need to be added.  Likewise, if braces were added to the outer \textit{if}
statement first, then this would shift the location of the inner \textit{if}
statement.  In other words, when refactorings affect nested language
constructs, offset/length computations can become complex.

So, manipulating source code using a mutable AST is simpler, and it more
directly expresses the programmer's intention.  Manipulating source code using
offset/length information leads to ugly client code.  The client code can
easily become littered with variables holding character offsets and methods
that perform intricate string manipulations.
